A Cost-based Approach for Fast Intrusion Detection

### 2017-06-05 tex/paper_v1.tex
final version

The aim of Intrusion Detection System(IDS) is to maximize detection accuracy as well as minimize corresponding costs. In this paper, I present a cost-based approach utilizing neural network with GPU acceleration for fast intrusion detection. In this approach, both computational cost and time cost are considered, and models are trained with respect to their different protocol types and services. Empirical experiments are carried out on off-line benchmark dataset NSL-KDD.

@@

\section{Introduction}
With the development of computing and Internet technology, Internet has became an important part of our daily life. However, this popularity has also brought security issues. 
Intrusion Detection Systems(IDSs) are designed to detect attacks, which help discover, determine, and identify unauthorized use, duplication, alteration, and destruction of information systems\cite{mukkamala2005cyber}. Specifically, IDSs are able to monitor intrusions and alert network administrators if necessary.

According to the detectable attacks, there are mainly three types of IDSs: misuse-based, anomaly-based, and hybrid\cite{buczak2016survey}.
Misuse-based IDSs, which are also called signature-based IDSs are designed to detect known attacks by comparing new packets and the signatures in the database.
Anomaly-based IDSs aims at modeling the normal network, and anomalies can be detected when happened. An attractive property of the anomaly-based IDSs is that they have the ability for detecting zero-day attacks.

Also with respect for using environment, IDSs can be used in real-time environment or off-line environment. Off-line IDSs have been studied extensively recently especially with the support of data mining and machine learning techniques, such as K-means\cite{lin2015cann} and ANN\cite{ingre2015performance} etc. And excellent detection results with high accuracies and low false alarm rates are demonstrated. For detailed introduction for these machine learning techniques, the survey\cite{buczak2016survey} is recommended.

However, these systems fail to provide real time information, which means they can only detect intrusions after they already happened for a while. However, it's obvious that as the complexity of features increase, the detection accuracy will also make an improvement. So a balance need to be maintained between accuracy and efficiency.

In this paper, the aim is to develop an IDS which can be utilized in real-time environment with high detection accuracy and low cost. I present a cost-based approach for anomaly intrusion detection. The traditional machine learning models are replaced by neural networks, which enable the computational cost to be minimized utilizing GPU acceleration. 
Also, the time cost is considered, and according to the time cost theory in \cite{lee2001real}, low level features are utilized with high priority. The benchmark dataset NSL-KDD is used for evaluation, which is a new version of dataset KDD Cup 1999\cite{cup2007available} and has significant improvements compared with KDD Cup 1999\cite{tavallaee2009detailed}.

The rest of the paper is organized as follows. Sec \ref{sec:relatedwork} gives a detailed review of current related work about intrusion detection. In Sec \ref{sec:methodology}, the new approach for intrusion detection is introduced. The experimental results are reported in Sec \ref{sec:experiment}. More future work is presented in Sec \ref{sec:futurework}. Finally, a conclusion is drew in Sec \ref{sec:conclusion}.

\section{Related Work}\label{sec:relatedwork}
\subsection{Detection Methodology}
Basically, there are three types of intrusion detection techniques: misuse-based, anomaly-based, and hybrid\cite{buczak2016survey}. Hybrid techniques are a combination for the two former techniques.

\subsubsection{Misuse-based Intrusion Detection}
The misuse-based intrusion detection is also known as signature-based or knowledge-based intrusion detection. A signature is a pattern or string that corresponds to a known attack or threat\cite{liao2013intrusion}. So in the misuse-based IDS, the process for coming packets is to compare them with the known signatures for recognizing possible intrusions.

The advantages of misuse-based method is that the misuse-based method is effective for detecting known type of attacks free from false alarms. However, the disadvantages are that it's usually hard to keep the signature database up to date. Also, the technique is incapable in detecting zero-day attacks or any attacks whose signatures are not included in the database.

\subsubsection{Anomaly-based Intrusion Detection}
An anomaly is a deviation to a known behavior, and profiles represent the normal or expected behaviors derived from monitoring regular activities, network connections, hosts or users over a period of time\cite{liao2013intrusion}. The anomaly-based techniques aims at modeling normal network status, and new packets are compared with the normal normal status to distinguish an attack.

The advantages of anomaly-based method is that they have the ability to recognize zero-day attacks. Another advantage is that unlike the misuse-based method, the profiles of normal activity are customized for every system, application, or network, thereby making it difficult for attackers to know which activities they can carry out undetected\cite{buczak2016survey}. The main disadvantage of anomaly-based techniques is the it may present high false alarm rates, since every behavior that has not happened before has the potential risk to be considered as a intrusion.

\subsubsection{Hybrid Intrusion Detection}
Hybrid techniques combine the former two types of methods.
They are implemented aiming at decrease the false alarm rate as well as detecting zero-day attacks.

\subsection{Feature Selection}
The reduction of features and feature selection are very important in increasing the efficiency of the detection.

This work\cite{chebrolu2005feature} aims at reduce the features used for misuse-based techniques, since the comparison between the whole packet and all the signatures can be redundant. And useful feature selection algorithms are proposed.
Also, the features on one specific dataset KDD Cup 1999 are analyzed by comparing the relevance between each features and information gain is employed to find the most discriminating features for each class\cite{kayacik2005selecting}.

Unsupervised machine learning techniques such as PCA\cite{liu2007hierarchical}\cite{lakhina2010feature} and K-means\cite{kumar2013k} are employed for feature selection. And all of these methods demonstrates good performance in feature selection.

However, these feature selection method ignore the time cost brought by the extraction process, so the selected features may bring severe time delay. In the cost-based approach presented in this paper, the cost factors are considerer, and a balance between cost and accuracy is maintained.

\subsection{Machine Learning Approaches}
In this section present a brief introduce of the utilization of machine learning techniques in intrusion detection. Specific machine learning algorithm is beyond the scope of this paper, and this book\cite{bishop2006pattern} is highly recommended for more information.

\subsubsection{Supervised Learning}
Supervised learning can be divided into generative models and deterministic models, and both contain plenty of proposed models. Some of the popular supervised learning methods are listed as follows: Artificial Neural Network, Bayesian Network, Decision Tree, Genetic Algorithms, Hidden Markov Models, Naive Bayes, Support Vector Machine etc.

Nearly all the former techniques in supervised learning are used in intrusion detection\cite{ingre2015performance}\cite{sommer2010outside}\cite{qiao2002anomaly}. Both the computational complexity and concrete introduction of the former techniques can be found in this work\cite{buczak2016survey}.

\subsubsection{Unsupervised Learning}
One of the most frequently used techniques is \textbf{clustering}.  
It is an unsupervised learning aims at clustering the data with some same property together. The main advantage of clustering is that it makes the labeling work not necessary. Specifically, for intrusion detection, the system can learn from audit data without requiring the network administrator to provide the attacks information.

There are many clustering method that demonstrates an excellent performance in detection intrusion experiments\cite{hendry2008intrusion}\cite{lin2015cann}\cite{kumar2013k}\cite{tsai2010triangle}. K-means, K-NN and etc are popular unsupervised machine learning approaches. Also, some work combine them together and better results are demonstrated\cite{lin2015cann}. However, as mentioned in the mentioned work, feature selection and reduction are urgently needed in methods like K-NN, because computing the similarities between high dimensions of features are expensive.

\subsubsection{Ensemble Learning}
Actually, the ensemble learning approaches do not belong to supervised or unsupervised learning.
Usually, the idea of ensemble learning is to use multiple weak models to build a strong model, since some classification tasks may be difficult for one classifier. And complex models with lots of parameters always tend to over fit on the training data.
Bagging and boosting are frequently used approaches in ensemble learning, \cite{dietterich2000experimental} gives a good example for utilizing ensemble learning methods in decision tree models.

\textbf{Bagging}\cite{breiman1996bagging} predictors is a method for generating multiple versions of a single model and using these to get an aggregated predictor. The aggregation averages over all the versions of models when predicting a result by majority voting. 
\textbf{Boosting}\cite{freund1995desicion} basically concentrates more the mis-classified data in the modeling process. Adaptive Boosting (AdaBoost)\cite{polikar2006ensemble} is one of the most popular algorithms in boosting. 
The \textbf{Random Forest}\cite{breiman2001random} another method that combines the decision trees and ensemble learning. The so-called forest is composed of many decision trees.

Ensemble learning methods are extensively utilized in intrusion detection\cite{zhang2008random}\cite{mukkamala2005intrusion}. The single model can be nearly all kinds of previous mentioned machine learning models.

@@

In the approach proposed in this paper, both the supervised learning and ensemble learning methods are employed. The combination of decision tree model and neural networks model demonstrates a great improvement compared with using single model for prediction in NSL-KDD dataset.

@@

\subsubsection{DARPA 1998 and DARPA 1999 Datasets}
One of the most commonly used datasets are the Defense Advanced Research Projects Agency (DARPA)
1998 and DARPA 1999 datasets\cite{lippmann2000evaluating}\cite{lippmann20001999}. Both of the datasets are created by the Cyber Systems and Technology Group of the Massachusetts Institute of Technology Lincoln Laboratory (MIT/LL). The datasets contain both network and operating system data. Totally, seven weeks of training data and two weeks of test data are collected, and attack simulations are organized during both of the periods.

\subsubsection{KDD Cup 1999 Dataset}
KDD Cup 1999 dataset\cite{tavallaee2009detailed} is one of the most widely used intrusion detection datasets, which is created for the KDD Cup challenge in 1999. The dataset is based on DARPA 1998 TCP/IP dataset and some basic features are extracted, as well as some more advanced features like some content features within a connection and traffic features computed using a two-second time window. Totally, $41$ features are extracted for a single connection.

\subsubsection{NSL-KDD Dataset}
As suggested in \cite{tavallaee2009detailed}, the original KDD Cup 1999 dataset has two important issues which highly affects the performance of evaluated systems, and results in a very poor evaluation of anomaly detection approaches. The NSL-KDD dataset consists of selected records of the complete KDD data set and free from the mentioned shortcomings in \cite{tavallaee2009detailed}.

\subsubsection{ADFA Intrusion Detection Datasets}
The ADFA Intrusion Detection Datasets cover both Linux and Windows\cite{creech2013generation}. Different from former introduced datasets, ADFA contains raw system call traces. 
A new host-based anomaly intrusion detection methodology is evaluated on ADFD\cite{creech2014semantic}, which utilizing discontiguous system call patterns to increase detection rates whilst reducing false alarm rates. Also, excellent results are demonstrated.

@@

As shown in Fig \ref{fig:overview}, a multi-model detector is presented here. Based on a decision tree structure, different models are used for every kind of packets with different protocol types and services.
And neural network models are trained for each level of features respectively.

\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.55]{img/overview-new}
\par\end{centering}
\caption{\label{fig:overview} An overview of the detection framework including the decision tree(top part) and the neural networks(bottom part). The total number of models are calculated on NSL-KDD dataset.}
\end{figure*}

@@

\subsubsection{Computational Cost}
One of the major challenges in real-time IDSs is the computational cost attached with the system. For example, the well-known K-nearest Neighbor Classifier\cite{liao2002use} performs well on off-line datasets, but it is not an appropriate detector in real time. Because the
KNNC calculates the similarity between the new 
packet and each packet in the training set to decide the label of the new packet. So KNNC is computationally
expensive when the incoming number of packets is huge, causing the system to be overloaded.

Traditionally, less features numbers and models with less parameters will lead to low computational cost. However, this may also result in lower detection accuracy or higher false alarm rate. Different from former approaches, utilizing deep neural networks, the computing tasks can be accelerated on GPU\cite{chetlur2014cudnn}. Also, the ability to represent any non-linear function in very few hidden layers and hidden neurons make the computational expense even less.

@@

Notice that the Level 3 features require iteration from the entire former two-second packet set, the time cost is extremely high compares with the lower level of features. Also, all the Level 3 features for connections coming at the same time can also be computed at the same time.

The process of utilizing different levels of features is:
\begin{itemize}
\item The lower level features are first used;
\item If the confidence of the classification is less than a threshold $\theta$, higher level features are extracted and are used combined with other lower level features;
\item If the extracted feature level is already $3$, then the eventual detection result is made without confidence calculating.
\end{itemize}
For example, both Level 1 and Level 2 features are considered after extracting the Level 2 features.

@@

Packets with different protocol types and different services tend to have different signatures to be malicious.
Thus different detectors should be trained, and for each new packet, its protocol type and service decides which detector to use. This model selection framework can be modeled by a decision tree.

As shown in Fig \ref{fig:overview}, the decision tree has very simple architecture in the proposed approach.

@@

Neural network models are at the bottom layer for the detection process, they trained for each protocol type and service respectively. Actually due to the less complexity for packets detection, neural network models perform nearly the same as traditional classifiers. The main reason of utilizing neural network models in this paper is the excellent GPU acceleration support for deep learning libraries, which will reduce the computational cost in some extent.

Neural network models with only one hidden layer is utilized in this paper, and the following experiments proves their capability for intrusion detection with such simple architectures.

@@

\begin{itemize}
\item Detection accuracy is simply calculated by comparing the predicted results with the ground truth labels.
\item The false alarm rate(FAR) is defined as Eq \ref{eq:false-alarm}, in which $FP, TN$ represent false positive and true negative samples respectively.

@@

\section{Experiments}\label{sec:experiment}
\subsection{Dataset}
As mentioned in Sec \ref{sec:publicdatasets}, NSL-KDD dataset consists of the selected records of the KDD Cup 1999 dataset, which is more challenging to detect compared with the original widely used benchmark KDD Cup 1999 dataset. 
The NLS-KDD dataset solves the issues of models trained on KDD Cup 1999 dataset, which have high seemingly accuracy but actually poor capability of anomaly detection.

@@

\subsection{Parameter Setting}
The data preprocessing method z-normalization is employed.
All the code for the neural network models is implemented in Python utilizing deep learning framework Tensorflow, and the neural network models are trained on one GPU GeForce GTX TITAN X for $5000$ epochs with early stop policy. All the neural networks consist of $1$ hidden layer, and the number of hidden neurons are set specifically for every model. The initial learning rate is $\eta = 0.0001$ and dropout ratio is $0.75$ to avoid over fitting. Finally, a softmax layer is utilized to calculate loss and Adam\cite{kingma2014adam} is employed to accelerate learning. The prediction results of the models are the values of the two output neurons.

\subsection{Baseline}
Support Vector Machine(SVM) is employed as a baseline method for each model respectively. The code are implemented in Python on Scikit-learn Library. And the following baseline results are evaluated on the same server with the proposed approach without GPU support for SVM method.

@@

\subsubsection{Detection Accuracy}
Tab \ref{tab:detection-accuracy} demonstrates the detection accuracy for each level of features. And apart from the over all accuracy, I take three representative protocol type and service(TCP finger/UDP domain\_u/ICMP eco\_i) as examples.

Observed from Tab \ref{tab:detection-accuracy}, we can see that SVM models and neural network models perform nearly the same on most of the detections. But neural network models demonstrate a good property on the three examples that as the level of features get higher, the accuracies never decay; while the SVM method present less stability. Actually, this property is quite important in the proposed approach, because the accuracy decay on higher feature level will result in some eventual detection results to be wrong. Since the results for packets that cannot be classified confidently by the lower feature level models are decided finally by the higher feature level models. Also, lots of time cost will be wasted for no improvement in detection accuracy.

Also, we can observe that the decision tree idea that to detect different protocol types and service with different models is quite reasonable. In Tab \ref{tab:detection-accuracy}, huge accuracy improvement has been made thanks to this separation.

@@

\subsubsection{Cost Analysis.}
Tab \ref{tab:cost} demonstrates the \textbf{computational cost} for each level of features. The same as former experiments, three representative protocol type and service are selected. Notice here I report the average time for computing single one packet feature item.

Thanks to GPU acceleration, it can be noticed that significant improvement have been made on computational cost utilizing neural network models, which is quite important in real time detection for the IDSs. Also the number of neurons in the hidden layer is reduced as small as possible, which also save the computational cost.

Also, \textbf{time cost} is considered. Every new packet are first detected according to their Level 1 features. If the confidence of the classification, which is the larger value of the two output neurons is less than $\theta=0.8$, than the higher features are extracted and higher level models are used. 
In the worst case, the packet will eventually get its label through Level 3 models. And more additional computational cost will also be spent.

The total cost is the sum of computational cost and time cost. The proposed approach is much more efficient compared with simply using Level 3 features for all the packets, and there is almost no decay of detection accuracy.

>>>>

The aim of real time Intrusion Detection Systems(IDSs) is to maximize detection accuracy as well as minimize corresponding costs. In this paper, I present a cost-based approach utilizing decision tree and neural network models with GPU acceleration for fast intrusion detection. In this approach, both computational cost and time cost are considered, and different models are trained with respect to the types of protocols and services. Empirical experiments are carried out on off-line benchmark dataset NSL-KDD.

@@

\section{Introduction}
With the development of computing and network technology, Internet has became an important part of our daily life. However, this popularity has also brought many security issues. 
Intrusion Detection Systems(IDSs) are designed to detect attacks. Specifically, it helps to discover, determine, and identify unauthorized use, duplication, alteration, and destruction of information systems\cite{mukkamala2005cyber}. Additionally, IDSs are able to monitor the system and alert network administrators if necessary.

With respect to the types of detectable attacks, there are mainly three types of IDSs: misuse-based, anomaly-based, and hybrid\cite{buczak2016survey} IDS.
Misuse-based IDSs, which are also called signature-based IDSs are designed to detect known attacks by comparing new packets and the predefined signatures stored in the database.
Anomaly-based IDSs aims at modeling the normal network, and anomalies can be detected when happened. Hybrid IDSs combine the former two techniques together.

Also with respect to the using environment, IDSs can be divided as the ones used in real-time environment or used in off-line environment. Off-line IDSs have been studied extensively recently especially based on data mining and machine learning techniques, such as K-means\cite{lin2015cann} and ANN\cite{ingre2015performance} etc. And excellent detection results with high accuracies and low false alarm rates on the off-line datasets are accomplished.

However, these IDSs always fail to provide real-time detection, which means they can only detect intrusions after the they already happened for a while and are not capable in defending the host from an attack at all. This is because the excellent off-line performance is often achieved thanks to the comprehensive extracted features and long time complicated computing. So a balance need to be maintained between accuracy and efficiency in real-time IDSs.

In this paper, the aim is to develop an IDS which can be utilized in real-time environment with high detection accuracy and low cost. For the minimization of computational cost, I present a cost-based approach for fast intrusion detection. The traditional machine learning models are replaced by neural networks, which enable the utilizing of GPU acceleration. 
Also, the time cost is considered, and according to the time cost theory in \cite{lee2001real}, low level features are utilized with high priority. The benchmark dataset NSL-KDD is used as testbed for evaluation, which is a new version of dataset KDD Cup 1999 dataset\cite{cup2007available} and has significant improvements compared with KDD Cup 1999 dataset\cite{tavallaee2009detailed}.

The rest of the paper is organized as follows. Sec \ref{sec:relatedwork} gives a detailed review of current related work about intrusion detection. In Sec \ref{sec:methodology}, the new approach for intrusion detection is proposed. The experimental results are reported in Sec \ref{sec:experiment}. More future work is presented in Sec \ref{sec:futurework}. Finally, a conclusion is drew in Sec \ref{sec:conclusion}.

\section{Related Work}\label{sec:relatedwork}
\subsection{Intrusion Types}
There are many types of intrusions.
According to KDD Cup 1999 dataset\cite{tavallaee2009detailed}, all the simulated attacks are classified into four categories:

\begin{itemize}
\item \textbf{Denial of Service Attack (DoS):} This type of attacks aim at making the services of target computer unavailable to its legitimate users by increasing the number of significant resource demanding processes to exhaust the target's network bandwidth , CPU load, or memory etc\cite{bhuyan2014detecting}. However, with the development of detection and IP tracking techniques, old single source attacks can be easily detected and the source can be rebuffed or shut down. Many kinds of  distributed Dos(DDos) attack techniques have been developed. One of the effective approaches to detect DDos attacks is to capture and recognize patterns of the DDos affected traffic\cite{robinson2015ranking}. This survey\cite{zargar2013survey} is recommended for more information.
\item \textbf{User to Root Attack (U2R)} This type of attacks refer to the behavior that the attacker tries to gain root access of the system. Usually the attacker enters the system as a normal user first and try to exploit some vulnerabilities in the system to gain illegal root access.
\item \textbf{Remote to Local Attack (R2L):} This type of attacks occur when an attacker tries to gain local access as a legitimate user on a remote server where actually he or she has no legitimate accounts. R2L attacks are often carried out by sending packets over network to the remote server\cite{mukkamala2005intrusion}. Usually, this illegal accesses are gained by taking advantages of vulnerabilities on the server.
\item \textbf{Probing Attack:} This type of attacks are basically an information collection behavior. The attacker scans the computers in a specific network to gather information or to exploit vulnerabilities.
For more information about IDSs specially designed for probing attacks, this work \cite{zargar2009identification} is recommended.
\end{itemize}

\subsection{Detection Methodology}
Basically, there are three types of intrusion detection techniques: misuse-based, anomaly-based, and hybrid\cite{buczak2016survey}. In the following parts, a brief introduction is presented.

\subsubsection{Misuse-based Intrusion Detection}
The misuse-based intrusion detection is also known as signature-based or knowledge-based intrusion detection. A signature is a pattern or string that corresponds to a known attack or threat\cite{liao2013intrusion}. So in the misuse-based IDS, the detection procedure for new coming packets is to compare them with the known signatures for recognizing possible intrusions.

The main advantage of misuse-based methods is that the method is effective for detecting known type of attacks and the detection results are free from false alarms. However, there are also some disadvantages. Firstly, it's usually hard to keep the signature database up to date. Secondly, this type of IDSs are incapable in detecting zero-day attacks or any attacks whose signatures are not included in the database, so their abilities are quite limited.

\subsubsection{Anomaly-based Intrusion Detection}
An anomaly is a deviation from a known behavior\cite{liao2013intrusion}. The anomaly-based techniques aim at modeling normal status including regular activities, normal network connections, hosts or users etc. And new packets different from normal status can be detected, the anomalies are considered to be an attack.

One of the most attracting advantages of anomaly-based methods is that they have the abilities to recognize zero-day attacks. Another advantage is that unlike the misuse-based methods, the profiles of normal activities are customized for every system, application, or network, thereby making it difficult for attackers to know which activities they can carry out undetected\cite{buczak2016survey}. The main disadvantage of anomaly-based techniques is the they may present high false alarm rates, since every behavior that has not happened before has the potential risk to be considered as an intrusion by this type of IDSs.

\subsubsection{Hybrid Intrusion Detection}
Hybrid techniques combine the former two types of techniques.
They are implemented aiming at decreasing the false alarm rate as well as detecting zero-day attacks. Actually, pure anomaly-based IDSs are rarely used.

\subsection{Feature Selection}\label{sec:feature-selection}
The reduction of features and feature selection are very important issues in increasing the efficiency of the detection.

This work\cite{chebrolu2005feature} aims at reducing the features used for misuse-based techniques, since the comparison between the whole packet and all the signatures can be time demanding and actually redundant. Also, useful feature selection algorithms are proposed in the paper.
Furthermore, the features on one specific dataset KDD Cup 1999 dataset are analyzed by comparing the relevance between each features and information gain is employed to find the most discriminating features for each class\cite{kayacik2005selecting}.

Unsupervised machine learning techniques such as PCA\cite{liu2007hierarchical}\cite{lakhina2010feature} and K-means\cite{kumar2013k} are employed for feature selection process. And all of these methods demonstrates good performance on reducing the dimension of features and maintaining only the most representative ones.

However, as these feature selection methods are nearly all used in off-line detection, they ignore the time cost brought by the feature extraction process. So the selected features may bring severe time delay. In the cost-based approach presented in this paper, the cost factors are considerer, and a balance between cost and accuracy is maintained.

\subsection{Machine Learning Approaches}
In this section, I present a brief introduce of the utilization of machine learning techniques in intrusion detection. Specific machine learning algorithms are beyond the scope of this paper, and this book\cite{bishop2006pattern} is highly recommended for more knowledge on pattern recognition machine learning.

\subsubsection{Supervised Learning}
Supervised learning can be divided into generative models and deterministic models, and both classes contain plenty of proposed models. Some of the popular supervised learning methods are listed as follows: Artificial Neural Network, Bayesian Network, Decision Tree, Genetic Algorithms, Hidden Markov Models, Naive Bayes, Support Vector Machine etc.

Nearly all the mentioned techniques in supervised learning are used in intrusion detection, only a few papers are listed here\cite{ingre2015performance}\cite{sommer2010outside}\cite{qiao2002anomaly}. Comprehensive computational complexity analysis and concrete introduction of the supervised machine learning techniques can be found in \cite{buczak2016survey} and \cite{tsai2009intrusion}.

\subsubsection{Unsupervised Learning}
One of the most frequently used unsupervised learning techniques is \textbf{clustering}.  
This method aims at clustering the data with some similar properties together. The main advantage of clustering is that it makes the labeling work not necessary. Specifically, for intrusion detection task, this type of IDS can learn from audit data without labeling or requiring the network administrator to provide concrete attacks' information.

There are many clustering methods that demonstrate excellent performance in the evaluation\cite{hendry2008intrusion}\cite{lin2015cann}\cite{kumar2013k}\cite{tsai2010triangle}. K-means, K-NN are the most frequently used unsupervised learning approaches. Also, some work combine K-means and K-NN together and better results are achieved\cite{lin2015cann}. However, as mentioned in Sec \ref{sec:feature-selection}, feature selection and reduction are urgently needed in methods like K-NN, because computing the similarities between high dimensional features are really expensive.

\subsubsection{Ensemble Learning}
Actually, the ensemble learning approaches do not belong to supervised or unsupervised learning.
Basically, the idea of ensemble learning is to use multiple weak models to build a strong model. This is because trying to solve some complicated tasks with a single model may lead the model to be extremely complex. 
And it is known that complex models with lots of parameters usually tend to over fit on the training data.
Bagging and boosting are frequently used approaches in ensemble learning, this work\cite{dietterich2000experimental} gives a good example for utilizing ensemble learning methods in decision tree models.

\textbf{Bagging}\cite{breiman1996bagging} is a method of generating multiple versions of a single model and using them to get an aggregated model. The aggregation averages over all the versions of models when predicting a result by majority voting. 
\textbf{Boosting}\cite{freund1995desicion} is based on the idea to concentrate more the mis-classified data of every epoch in the training process. Adaptive Boosting (AdaBoost)\cite{polikar2006ensemble} is one of the most frequently used algorithms in boosting. 
The \textbf{Random Forest}\cite{breiman2001random} is a method that combines the decision trees and ensemble learning. The so-called forest is composed of many decision trees. Also, bagging techniques are employed to avoid over fitting. The random forest has been proved to be effective in many tasks.

Ensemble learning methods are extensively utilized in intrusion detection\cite{zhang2008random}\cite{mukkamala2005intrusion}. The type of single models used for the aggregation can be nearly all kinds of previous mentioned machine learning models.

@@

In the approach proposed in this paper, both the supervised learning and ensemble learning methods are employed. The combination of decision tree model and neural networks model demonstrates a great improvement compared with using single model for prediction in NSL-KDD dataset. For example, random forest is used both in mis-use based IDS\cite{gharibian2007comparative} and anomaly-based IDS\cite{zhang2008random}.

@@

\subsubsection{DARPA 1998 and DARPA 1999 Datasets}
One of the most commonly used datasets are the Defense Advanced Research Projects Agency (DARPA)
1998 and DARPA 1999 datasets\cite{lippmann2000evaluating}\cite{lippmann20001999}. Both of the datasets are created by the Cyber Systems and Technology Group of the Massachusetts Institute of Technology Lincoln Laboratory (MIT/LL). The datasets contain both network and operating system data. Totally, seven weeks of training data and two weeks of test data are collected, and attack simulations are organized during both of the periods. Notice that for anomaly detection, there are some types of attacks only appear in the test data.

\subsubsection{KDD Cup 1999 Dataset}
KDD Cup 1999 dataset\cite{tavallaee2009detailed} is also one of the most widely used intrusion detection datasets, which is created for the KDD Cup challenge in 1999. The dataset is based on DARPA 1998 TCP/IP dataset and some basic features are extracted. Also some more advanced features like some content features within a connection and traffic features computed using a two-second time window are computed. Totally, $41$ features are extracted for a single connection.

\subsubsection{NSL-KDD Dataset}
As suggested in \cite{tavallaee2009detailed}, the original KDD Cup 1999 dataset has two important issues which highly affects the performance of evaluated systems, and results in a very poor evaluation of anomaly detection approaches.The NSL-KDD dataset consists of only the selected records of the complete KDD Cup 1999 dataset and free from the mentioned shortcomings in \cite{tavallaee2009detailed}.

\subsubsection{ADFA Intrusion Detection Datasets}
The ADFA Intrusion Detection Datasets contain data both generated from Linux and Windows\cite{creech2013generation} systems. Different from former introduced datasets, ADFA contains raw system call traces. 
A new host-based anomaly intrusion detection methodology is evaluated on ADFD\cite{creech2014semantic}, which utilizing discontiguous system call patterns to increase detection rates whilst reducing false alarm rates.

@@

As shown in Fig \ref{fig:overview}, a multi-model detector is presented here. Different neural network models are used for packets with different protocol types and services based on a decision tree model.
Additionally, neural network models are trained for each level of features respectively.

\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.55]{img/overview-new}
\par\end{centering}
\caption{\label{fig:overview} An overview of the detection framework including the decision tree model(top part) and the neural network models(bottom part). The sequence number of models are counted based on NSL-KDD dataset.}
\end{figure*}

@@

\subsubsection{Computational Cost}
One of the major challenges for real-time IDSs is the computational cost attached with the detection procedure. For example, the well-known K-nearest Neighbor Classifier(KNNC)\cite{liao2002use} performs well on off-line datasets, but it is not an appropriate detector in real time. Because the
KNNC calculates the similarity between the new 
packet and every packet in the training set to decide the label of the new packet. So KNNC is computationally
expensive when the incoming number of packets is huge, and the system may be overloaded.

Traditionally, reducing the dimension of features and number of parameters in detection models will lead to lower computational cost. However, this may also result in lower detection accuracy or higher false alarm rate. Different from former approaches, in utilizing deep neural networks, the computing tasks can be accelerated on GPU\cite{chetlur2014cudnn}. Also, the ability of neural network models to represent any non-linear function in very few hidden layers and hidden neurons makes the computational expense even less.

@@

Notice that the Level 3 features require iteration from the entire former two-second packet set, the time cost is extremely high compared with the lower level of features. Also, all the Level 3 features for connections coming at the same time can also be computed at the same time.

The process of utilizing different levels of features is:
\begin{itemize}
\item The lower level features are first used;
\item If the confidence of the classification is less than a threshold $\theta$, higher level features are extracted and are used combined with other lower level features;
\item If the extracted feature level has already reached $3$, the eventual detection result is directly made.
\end{itemize}
For example, both Level 1 and Level 2 features are considered after extracting the Level 2 features.

@@

Packets with different protocol types and different services tend to have different signatures to be malicious.
Thus different detectors should be trained respectively. For each new packet, its protocol type and service decides which detector to use. This model selection framework is be modeled by a decision tree.

As shown in Fig \ref{fig:overview}, the decision tree has very intuitive architecture in the proposed approach.

@@

Neural network models are at the leaf nodes for the detection tree, they are trained for each protocol type and service respectively. Actually due to the less complexity for packets detection, neural network models perform nearly the same as traditional classifiers. The main reason of utilizing neural network models in this paper is the excellent GPU acceleration support for deep learning libraries, which will reduce the computational cost to some extent.

Neural network models with only one hidden layer is utilized in this paper, and the following experiments proves their capability for intrusion detection even if they only have such simple architectures.

@@

\begin{itemize}
\item Detection accuracy is calculated by comparing the predicted results with the ground truth labels.
\item The false alarm rate(FAR) is defined as Eq \ref{eq:false-alarm}, in which $FP, TN$ represent false positive and true negative samples respectively.

@@

\section{Experiments}\label{sec:experiment}
\subsection{Dataset}
As mentioned in Sec \ref{sec:publicdatasets}, NSL-KDD dataset consists of the selected records of the KDD Cup 1999 dataset, which makes it more challenging to be detected compared with the original widely used benchmark KDD Cup 1999 dataset. 
The NLS-KDD dataset solves the issues of models trained on KDD Cup 1999 dataset, which have high seemingly accuracy but actually poor capability of anomaly detection.

@@

\subsection{Parameter Setting}
The data preprocessing method z-normalization is employed.
All the code for training the neural network models is implemented in Python utilizing deep learning framework Tensorflow, and the neural network models are trained on one GPU GeForce GTX TITAN X for $5000$ epochs with early stop policy. All the neural network models consist of $1$ hidden layer, and the number of hidden neurons are set specifically for every model. The initial learning rate is $\eta = 0.0001$ and dropout ratio is set to be $0.75$ to avoid over fitting. Eventually, a softmax layer is utilized to calculate loss and Adam\cite{kingma2014adam} is employed to accelerate learning. The final prediction results of the models are the values of the two output neurons.

\subsection{Baseline}
Support Vector Machine(SVM) is employed as a baseline method for each model respectively. The code are implemented in Python utilizing Scikit-learn Library. And the following baseline results are evaluated on the same server with the proposed approach without GPU support for SVM method. Also, empirically, the SVM models are also combined with the same decision tree model in the proposed approach to accomplish a comparative result.

@@

\subsubsection{Detection Accuracy}
Tab \ref{tab:detection-accuracy} demonstrates the detection accuracy for each level of features. And apart from the over all accuracy, I take three representative protocol types and services(TCP finger/UDP domain\_u/ICMP eco\_i) as examples.

Observed from Tab \ref{tab:detection-accuracy}, we can see that SVM models and neural network models perform nearly the same on most of the detections. But neural network models demonstrate a good property on the three examples that as the level of features increases, the accuracies never decay; while the SVM method present less stability. Actually, this property is quite important in the proposed approach, because the accuracy decay on higher feature levels will result in some eventual detection results to be wrong. Since the results for packets that cannot be classified confidently by the lower feature level models are all finally decided by the Level 3 feature models. Also, lots of time cost will be wasted for no improvement in detection accuracy.

Also, we can observe that the idea of utilizing decision tree models to detect different protocol types and service with different models is quite reasonable. In Tab \ref{tab:detection-accuracy}, huge accuracy improvement has been made thanks to this separation.

@@

\subsubsection{Cost Analysis.}
Tab \ref{tab:cost} demonstrates the \textbf{computational cost} for each level of features. The same as former experiments, three representative protocol type and service are selected. Here I report the average time for computing single one packet feature item.

Thanks to GPU acceleration, it can be noticed that significant improvement has been made on computational cost utilizing neural network models, which is quite important in real time detection for the IDSs. Also practically, the number of neurons in the hidden layer is reduced as small as possible for each model, which also save the computational cost.

Maybe you could notice that in the SVM method, the computational cost may even decay as the number of feature dimensions increase. This is because in the SVM model, only the training data on the decision boundary is used in the calculation for prediction, so the actual computational cost does not only depend on the dimension of features. For more theoretical knowledge of SVM, this book\cite{bishop2006pattern} is recommended.

Also, \textbf{time cost} is considered. Following the process of utilizing different levels of features, new packet is first detected according to its Level 1 features. If the confidence of the classification, which is the larger value of the two output neurons is less than $\theta=0.8$, than the higher features are extracted and higher level models are used. 
In the worst case, the packet will eventually get its label through Level 3 models. And more additional computational cost will also be spent at the same time.

The total cost is the sum of computational cost and time cost. The proposed approach is much more efficient compared with simply using Level 3 features for all the packets. Additionally, there is almost no decay of detection accuracies or increment of false alarm rates.
